Reconstructing Propositional Proofs in Type Theory

### 2017-10-25 paper/sections/acknowledgements.tex
more fixes.

Last but not least, I thankfully acknowledge
Andreas Abel and Chalmers University of Goteborg, Sweden for
inviting me to be part of the \Agda Implementors’ Meeting XXV.
I was honored to present part of this research in the meeting.

>>>>

Last but not least, I thankfully acknowledge
Andreas Abel and Chalmers University of Goteborg, Sweden for
inviting me to be part of the \Agda Implementors’ Meeting XXV
where I presented part of this research in the meeting.

### 2017-10-25 paper/sections/proof-reconstruction/resolve.tex
more fixes.

\begin{myremark}
In \TPTP syntax, the formulas are in left-associative form by default.
Despite of that convention, \Metis understands a formula to be in right-associative form by default. This is a matter to take into account for the proof-reconstruction.
\end{myremark}

@@

\begin{mainlemma}
\label{lem:build-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fbuild_{∨}~φ~ψ$, where,
\begin{equation}
  \begin{split}
  \label{eq:build-or}
  &\fbuild_{∨} : \Prop → \Prop → \Prop\\
  &\fbuild_{∨}~φ~ψ~ =%\\
  % &\hspace{3mm}
  \begin{cases}
  ψ, &\text{ if } φ ≡ ψ\\
  ψ, &\text{ if } ψ ≡ ψ₁ ∨ ψ₂\text{, } ψᵢ ≡ \fbuild_{∨}~φ~ψᵢ\text{ for some }i=1, 2\\
  φ, &\text{ otherwise.}
  \end{cases}
  \end{split}
\end{equation}
\end{mainlemma}

From now, we assume all propositions to be right-associative unless
otherwise stated.
The $\ffactor$ function in~(\ref{eq:factor-definition})
simplifies a special case of disjunction, the repeated disjuncts
(\eg, $\ffactor~(φ ∨ φ)~=~φ$).
Notice that other cases like $φ~∨~(ψ~∨~φ)$ do no reduce to $ψ~∨~φ$.

\begin{mainlemma}
\label{lem:factor}
 If $Γ ⊢ φ$ then $Γ ⊢ \ffactor~φ$, where,
\begin{align}
\begin{split}
  \label{eq:factor-definition}
  &\ffactor : \Prop → \Prop\\
  &\ffactor~φ~ =
  \begin{cases}
    φ₁,  &\text{ if }φ ≡ φ₁ ∨ φ₂ \text{ and } φ₁ ≡ \ffactor~φ₂\\
    φ,   &\text{ otherwise.}
  \end{cases}
\end{split}
\end{align}
\end{mainlemma}

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
We define the $\fsbuild_{∨}$ function in~(\ref{eq:strong-build-or}).
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fsbuild_{∨}~φ~ψ$, where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\fsbuild_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{lll}
    \fsbuild_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor~(\fbuild_{∨}~φ₁~ψ ∨ \fbuild_{∨}~φ₂~ψ)\\
    \fsbuild_{∨}&φ~ψ &= \fbuild_{∨}~φ~ψ
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

\begin{myexample}
We build the disjunction $(p ∨ q) ∨ r$ from the disjunction
$r ∨ (q ∨ p)$ based on the idea of the Lemma~\ref{lem:sbuild-or}.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

  \AxiomC{$Γ, q ∨ p ⊢ (p ∨ q) ∨ r$}

@@

\begin{mainlemma}
  \label{lem:reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}~φ~ψ~$, where,
  \begin{align}
    \begin{split}
    \label{eq:reorder-or}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨}~φ~ψ~= \fsbuild_{∨}~(\fassoc_{∨}~φ)~ψ.
    \end{split}
  \end{align}
\end{mainlemma}

@@

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$, where,
    \begin{align}
      \begin{split}
      \label{eq:reorder-and}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                   &\text{ and } ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

\begin{myexample}\hspace{10cm}
\begin{enumerate}
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(r ∧ q ∧ p)~=~(r ∧ q ∧ p).$
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(r ∧ r ∧ p)~=~(r ∧ q ∧ p).$
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(k ∧ q ∧ p)~=~(p ∧ q ∧ r).$
  \item $\freorder_{∧}~((p ∨ q) ∧ r)~(r ∧ (q ∨ p))~=~((p ∨ q) ∧ r).$
\end{enumerate}
\end{myexample}

In the last example, Item~4, we could not build the conjunction
$r ∧ (q ∨ p)$ since $p ∨ q$ is not syntactical equal to $q ∨ p$.
We solve this issue in the Lemma~\ref{lem:reorder-and-or} by
using the $\fconjunct_{∨}$ function defined in~(\ref{eq:conjunct-or}).

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$, where,
  \begin{equation*}
    \begin{split}
    \label{eq:conjunct-or}
      &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
      &\fconjunct_{∨}~φ~ψ =
      \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
        ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
           &\text{ and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
        φ, &\text{ otherwise.}
      \end{cases}
    \end{split}
  \end{equation*}
\end{mainlemma}

Now, we are able to reorder conjunctive normal forms using
$\freorder_{∧∨}$ function defined in (\ref{eq:reorder-and-or})
that unifies the previous lemma showed above.

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align}
      \begin{split}
      \label{eq:reorder-and-or}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{ and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

Now, we are in a position to reconstruct \resolve rule. This is the
first application of the Lemma \ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem showed in
Fig.~\ref{fig:metis-inferences} in the \TSTP derivations.
This rule has three arguments: a positive literal $l$ to apply the
resolution theorem and two derivations.
The positive literal $l$ must occur in
the formula from the first derivation and the
negative literal $¬ l$ must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

\begin{myexamplenum}\hspace{10cm}
\label{ex:resolve-tstp}
\begin{verbatim}
cnf(ref4, (~ r | p | q), inf(canonicalize, [norm5])).
cnf(ref5, (p | q | r), inf(canonicalize, [norm6])).
cnf(ref6, (p | q), inf(resolve, [$cnf(r)], [ref5, ref4])).
\end{verbatim}
In the excerpt above, we apply resolution for the first two formulas,
$¬ r ∨ p ∨ q$ and $p ∨ q ∨ r$. The last line tells us the literal mentioned
for resolution is $r$. Syntactically speaking,
we can not derive neither the conclusion $p ∨ q$ in \verb!ref6! nor apply the resolution theorem with \verb!ref4! and \verb!ref5! since the formulas does not fit the pattern required.

If the scenario would have other like replacing \verb!ref5! by
\begin{verbatim}
cnf(ref5, (r | p | q), inf(canonicalize, [norm6])).
\end{verbatim}
We have could derive $(p ∨ q) ∨ (p ∨ q)$ but again, that is not the expected result.
\end{myexamplenum}

Therefore, we perform a sequence of rearrangements inside the
formulas involved to match with the pattern expected by the resolution theorem
described in Fig.~\ref{fig:metis-inferences}. Using reordering after
applying a convenient version of the resolution theorem defined
in~(\ref{eq:rsol}) we get the expect result.

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rsol}~φ$, where,
  \begin{equation}
    \label{eq:rsol}
    \begin{aligned}
    &\rm{rsol} : \Prop \to \Prop\\
    &\begin{array}{ll}
      \rm{rsol}~φ &=
        \begin{cases}
          φ₂,      &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₂)\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₄)\\\
          φ, &\text{ otherwise.}
        \end{cases}
      \end{array}
      \end{aligned}
\end{equation}
\end{mainlemma}

\begin{mainth}
  \label{thm:resolve}
  If $Γ ⊢ φ₁$, $Γ ⊢ φ₂$, a literal $l : \Prop$, and $ψ : \Prop$ then
  $Γ ⊢ \fresolve~φ₁~φ₂~l~ψ$, where,
  \begin{equation}
  \begin{split}
  \label{eq:resolve}
    &\fresolve : \Prop \to \Prop \to \Prop \to \Prop \to \Prop\\
    &\fresolve~φ₁~φ₂~l~ψ =
      \rm{rsol}~(\freorder_{∨}~φ₁~(l ∨ ψ) ∧ \freorder_{∨}~φ₂~(¬ l ∨ ψ)).
  \end{split}
  \end{equation}
\end{mainth}

\Needspace{3\baselineskip}
\begin{proof}
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ₁$}
    \RightLabel{Lemma~\ref{lem:reorder-and-or}}
    \UnaryInfC{$Γ ⊢ \freorder_{∨}~φ₁~(l ∨ ψ)$}
    \RightLabel{Lemma~\ref{lem:reorder-and-or}}
    \AxiomC{$Γ ⊢ φ₂$}
    \UnaryInfC{$Γ ⊢ \freorder_{∨}~φ₂~(¬ l ∨ ψ)$}
    \RightLabel{∧-intro}
    \BinaryInfC{$Γ ⊢ \freorder_{∨}~φ₁~(l ∨ ψ) ∧ \freorder_{∨}~φ₂~(¬ l ∨ ψ)$}
    \RightLabel{Lemma~\ref{lem:rsol}}
    \UnaryInfC{$Γ ⊢ \rm{rsol}~(\freorder_{∨}~φ₁~(l ∨ ψ) ∧
     \freorder_{∨}~φ₂~(¬ l ∨ ψ))$}
    \RightLabel{by (\ref{eq:resolve})}
    \UnaryInfC{$Γ ⊢ \fresolve~φ₁~φ₂~l~ψ$}
  \end{bprooftree}
  \end{equation*}
\end{proof}

\begin{myexample}
Continuing with the problem presented in the Example~\ref{ex:resolve-tstp},
we can use Theorem \ref{thm:resolve} to derive $Γ ⊢ p ∨ q$.
Let be $φ₁ ≡ p ∨ q ∨ r$ and $φ₂ ≡ ¬ r ∨ p ∨ q$.

>>>>

\begin{myremark}
In \TPTP syntax, the formulas are in left-associative form by default.
Despite of that convention, \Metis assumes the formulas to be in
right-associative form by default. This is a matter to take into account for
the proof-reconstruction.
\end{myremark}

@@

\begin{mainlemma}
\label{lem:build-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fbuild_{∨}~φ~ψ$ where,
\begin{equation}
  \begin{split}
  \label{eq:build-or}
  &\fbuild_{∨} : \Prop → \Prop → \Prop\\
  &\fbuild_{∨}~φ~ψ~ =%\\
  % &\hspace{3mm}
  \begin{cases}
  ψ, &\text{ if } φ ≡ ψ\\
  ψ, &\text{ if } ψ ≡ ψ₁ ∨ ψ₂\text{ and } ψᵢ ≡ \fbuild_{∨}~φ~ψᵢ\text{ for some }i=1, 2\\
  φ, &\text{ otherwise.}
  \end{cases}
  \end{split}
\end{equation}
\end{mainlemma}

From now, we assume all propositions to be right-associative unless
otherwise stated.

The $\ffactor$ function in~(\ref{eq:factor-definition})
simplifies a special case of disjunction, the repeated disjuncts
(\eg, $\ffactor~(φ ∨ φ)~=~φ$).
Notice that other cases like $φ~∨~(ψ~∨~φ)$ do no reduce to $ψ~∨~φ$.

\begin{mainlemma}
\label{lem:factor}
 If $Γ ⊢ φ$ then $Γ ⊢ \ffactor~φ$ where,
\begin{align}
\begin{split}
  \label{eq:factor-definition}
  &\ffactor : \Prop → \Prop\\
  &\ffactor~φ~ =
  \begin{cases}
    φ₁,  &\text{ if }φ ≡ φ₁ ∨ φ₂ \text{ and } φ₁ ≡ \ffactor~φ₂\\
    φ,   &\text{ otherwise.}
  \end{cases}
\end{split}
\end{align}
\end{mainlemma}

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
We define the $\fsbuild_{∨}$ function in~(\ref{eq:strong-build-or})
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fsbuild_{∨}~φ~ψ$ where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\fsbuild_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{lll}
    \fsbuild_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor~(\fbuild_{∨}~φ₁~ψ ∨ \fbuild_{∨}~φ₂~ψ)\\
    \fsbuild_{∨}&φ~ψ &= \fbuild_{∨}~φ~ψ
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

\begin{myexample}
We build the disjunction $(p ∨ q) ∨ r$ from the disjunction
$r ∨ (q ∨ p)$ using Lemma~\ref{lem:sbuild-or}.
\begin{equation*}
(\mathcal{D})\hspace{3mm}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{$Γ, q ∨ p ⊢ (p ∨ q) ∨ r$}

@@

\begin{mainlemma}
  \label{lem:reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}~φ~ψ~$ where,
  \begin{align}
    \begin{split}
    \label{eq:reorder-or}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨}~φ~ψ~= \fsbuild_{∨}~(\fassoc_{∨}~φ)~ψ.
    \end{split}
  \end{align}
\end{mainlemma}

@@

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$ where,
    \begin{align}
      \begin{split}
      \label{eq:reorder-and}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                   &\text{ and } ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

\begin{myexample}
\begin{equation}
\label{ex:reorder-example}
\begin{aligned}
\begin{array}{llll}
  \freorder_{∧} &(p ∧ q ∧ r)   &(r ∧ q ∧ p)     ~&=~(r ∧ q ∧ p),\\
  \freorder_{∧} &(p ∧ q ∧ r)   &(r ∧ r ∧ p)     ~&=~(r ∧ q ∧ p),\\
  \freorder_{∧} &(p ∧ q ∧ r)   &(k ∧ q ∧ p)     ~&=~(p ∧ q ∧ r),\\
  \freorder_{∧} &((p ∨ q) ∧ r) &((r ∧ (q ∨ p)) ~&=~((p ∨ q) ∧ r).
\end{array}
\end{aligned}
\end{equation}
\end{myexample}

In the last example in (\ref{ex:reorder-example}), we could not build the conjunction
$r ∧ (q ∨ p)$ since $p ∨ q$ is not syntactical equal to $q ∨ p$.
We solve this issue in the Lemma~\ref{lem:reorder-and-or} by
using the $\fconjunct_{∨}$ function defined in~(\ref{eq:conjunct-or}).

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$ where,
  \begin{equation*}
    \begin{split}
    \label{eq:conjunct-or}
      &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
      &\fconjunct_{∨}~φ~ψ =
      \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
        ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
           &\text{ and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
        φ, &\text{ otherwise.}
      \end{cases}
    \end{split}
  \end{equation*}
\end{mainlemma}

We are able now to reorder conjunctive normal forms using the
$\freorder_{∧∨}$ function defined in (\ref{eq:reorder-and-or})
by using the previous lemma.

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$ where,
   \begin{align}
      \begin{split}
      \label{eq:reorder-and-or}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{ and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

Now, we are ready to reconstruct the \resolve rule. This is the
first application of the Lemma \ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem.
This rule has three arguments: a positive literal $ℓ$to apply the
resolution theorem and two derivations.
Recall a\emph{literal} is an atom (positive literal) or a negation
of an atom (negative literal).
The positive literal $ℓ$ must occur in
the formula from the first derivation and the
negative literal $¬ ℓ$ must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

\begin{myexamplenum}\hspace{10cm}
\label{ex:resolve-tstp}
\begin{verbatim}
cnf(ref4, (¬ r ∨ p ∨ q), inf(canonicalize, [norm5])).
cnf(ref5, (p ∨ q ∨ r), inf(canonicalize, [norm6])).
cnf(ref6, (p ∨ q), inf(resolve, r, [ref5, ref4])).
\end{verbatim}
In the excerpt above, we apply resolution to the first two formulas,
$¬ r ∨ p ∨ q$ and $p ∨ q ∨ r$. The last line tells us the literal used
for resolution is $r$. Syntactically speaking,
we can not derive neither the conclusion $p ∨ q$ in \verb!ref6! nor apply the resolution theorem with \verb!ref4! and \verb!ref5! since the formulas do not fit the pattern required.

If the scenario would have other like replacing \verb!ref5! by
\begin{verbatim}
cnf(ref5, (r ∨ p ∨ q), inf(canonicalize, [norm6])).
\end{verbatim}
The \resolve rule have could derive $(p ∨ q) ∨ (p ∨ q)$, but again, that is not the expected result.
\end{myexamplenum}

Therefore, we perform a sequence of rearrangements inside the
formulas involved to match with the pattern expected by the resolution theorem
described in Fig.~\ref{fig:metis-inferences}. Using reordering after
applying a customized version of the resolution theorem defined
in~(\ref{eq:rsol}) we get the expect result.

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rsol}~φ$ where,
  \begin{equation}
    \label{eq:rsol}
    \begin{aligned}
    &\rm{rsol} : \Prop \to \Prop\\
    &\begin{array}{ll}
      \rm{rsol}~φ &=
        \begin{cases}
          φ₂,      &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₂)\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₄)\\\
          φ, &\text{ otherwise.}
        \end{cases}
      \end{array}
      \end{aligned}
\end{equation}
\end{mainlemma}

\begin{mainth}
  \label{thm:resolve}
  If $Γ ⊢ φ₁$, $Γ ⊢ φ₂$, let $ℓ$ be a literal $ℓ : \Prop$, and $ψ : \Prop$ then
  $Γ ⊢ \fresolve~φ₁~φ₂~ℓ~ψ$ where,
  \begin{equation}
  \begin{split}
  \label{eq:resolve}
    &\fresolve : \Prop \to \Prop \to \Prop \to \Prop \to \Prop\\
    &\fresolve~φ₁~φ₂~ℓ~ψ =
      \rm{rsol}~(\freorder_{∨}~φ₁~(ℓ ∨ ψ) ∧ \freorder_{∨}~φ₂~(¬ ℓ ∨ ψ)).
  \end{split}
  \end{equation}
\end{mainth}

\Needspace{3\baselineskip}
\begin{proof}
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ₁$}
    \RightLabel{Lemma~\ref{lem:reorder-and-or}}
    \UnaryInfC{$Γ ⊢ \freorder_{∨}~φ₁~(ℓ ∨ ψ)$}
    \RightLabel{Lemma~\ref{lem:reorder-and-or}}
    \AxiomC{$Γ ⊢ φ₂$}
    \UnaryInfC{$Γ ⊢ \freorder_{∨}~φ₂~(¬ ℓ ∨ ψ)$}
    \RightLabel{∧-intro}
    \BinaryInfC{$Γ ⊢ \freorder_{∨}~φ₁~(ℓ ∨ ψ) ∧ \freorder_{∨}~φ₂~(¬ ℓ ∨ ψ)$}
    \RightLabel{Lemma~\ref{lem:rsol}}
    \UnaryInfC{$Γ ⊢ \rm{rsol}~(\freorder_{∨}~φ₁~(ℓ ∨ ψ) ∧
     \freorder_{∨}~φ₂~(¬ ℓ ∨ ψ))$}
    \RightLabel{by (\ref{eq:resolve})}
    \UnaryInfC{$Γ ⊢ \fresolve~φ₁~φ₂~ℓ~ψ$}
  \end{bprooftree}
  \end{equation*}
\end{proof}

\begin{myexample}
Continuing with the problem presented in Example~\ref{ex:resolve-tstp},
we can use Theorem \ref{thm:resolve} to derive $Γ ⊢ p ∨ q$.
Let be $φ₁ ≡ p ∨ q ∨ r$ and $φ₂ ≡ ¬ r ∨ p ∨ q$.
